---
title: "BIPPS Simulation Example: Varying number of latent fators"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
    embed_resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  comment = "#>"
)
```

# Overview

This Rmarkdown demonstrates how to run Bayesian Inference for Point Patterns in Space (BIPPS) simulations with varying numbers of latent factors (k). We'll generate synthetic spatial point process data with a known number of latent factors, then fit models with different numbers of factors to evaluate model performance under different specifications.

## What is BIPPS?

BIPPS (Bayesian Inference for Point Patterns in Space) is a spatial statistical modeling approach for analyzing point patterns or count data across a spatial domain. It uses latent Gaussian processes to model the underlying intensity of a point process, making it useful for applications like ecological data, disease mapping, or astronomical observations.

## Purpose of This Analysis

In this example, we'll:

1. Generate synthetic spatial point process data with a known number of latent factors (k)
2. Fit BIPPS models using different numbers of latent factors
3. Compare model performance to determine how well we can recover the true structure

# Setup

First, let's load the necessary libraries and set up our environment:

```{r load_libraries}
# Load the BIPPS package
# If you're using a development version, uncomment:
devtools::load_all()

# Load other required packages
library(magrittr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(posterior)
library(bayesplot)
library(ggdist)
library(latex2exp)
library(patchwork)

# Set random seed for reproducibility
set.seed(24)
```

# Simulation Parameters

We'll define all parameters needed for our simulation:

```{r simulation_parameters}
# MCMC settings
n_samples <- 500   # Number of samples to collect
n_burnin <- 500    # Number of burn-in iterations
n_thin <- 1        # Thinning rate
n_threads <- 2     # Number of threads for parallel processing
chains <- 1        # Number of MCMC chains

# Simulation parameters
actual_k <- 3      # True number of latent factors
trial_ks <- c(2,3,4)  # Different k values to try in our models

# Spatial grid specifications
nx <- 30           # Number of grid points in x direction
ny <- 30           # Number of grid points in y direction
n <- nx * ny       # Total number of grid points
x_max <- 1919      # Maximum x coordinate
y_max <- 1439      # Maximum y coordinate

# Model parameters
block_size <- 50   # Block size for MCMC sampling
starting <- list(phi = 1)  # Starting values for parameters
prior <- list(phi = c(0.1, 10))  # Prior specifications
phi_range <- c(1, 3)  # Range for spatial decay parameter phi
sample_theta <- TRUE  # Whether to sample theta parameter
mu <- -1              # Mean parameter for the baseline intensity
sigmasq <- 1          # Variance parameter
p <- 1                # Number of covariates (intercept only in this case)
q <- 4                # Dimension of the response
num_images <- 3       # Number of simulated images/replicates

# Create output directories if needed
if (!dir.exists("output")) {
  dir.create("output")
}
if (!dir.exists("figures")) {
  dir.create("figures")
}
```

# Generate Synthetic Data

## Create Spatial Grid

First, we'll create a grid of spatial coordinates:

```{r create_grid}
# Create a regular grid of coordinates
coords <- expand.grid(
  x = seq(0, x_max, length.out = nx),
  y = seq(0, y_max, length.out = ny)
)

# Normalize coordinates to [0,1] range
coords <- coords / max(x_max, y_max)
c_mat <- as.matrix(coords)

# Calculate distance matrix between all coordinates
d_coords <- as.matrix(dist(c_mat))

# Visualize the spatial grid
ggplot(coords, aes(x = x, y = y)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Spatial Grid for Simulation",
    subtitle = paste0(nx, "×", ny, " grid (", n, " total points)")
  )
```

## Generate Latent Spatial Fields

Now we'll generate the latent spatial fields using Gaussian processes with different spatial decay parameters:

```{r generate_latent_fields}
# Sample decay parameters for each latent factor
philist <- runif(actual_k, phi_range[1], phi_range[2])
print(paste("Spatial decay parameters (phi):", paste(round(philist, 2), collapse = ", ")))

# Create Cholesky factors of the covariance matrices
# Each latent factor has its own spatial correlation structure
LClist <- 1:actual_k %>% lapply(\(i) t(chol(
  exp(- philist[i] * d_coords)
)))

# Generate latent spatial fields for multiple images
VV <- lapply(1:num_images, \(j) {
  # For each image, generate 'actual_k' latent spatial fields
  wlist <- lapply(1:actual_k, \(i) LClist[[i]] %*% rnorm(n))
  
  # Combine all latent fields into a matrix
  do.call(cbind, wlist)
})

# Visualize first latent field from the first image
if (requireNamespace("fields", quietly = TRUE)) {
  fields::image.plot(
    seq(0, x_max, length.out = nx),
    seq(0, y_max, length.out = ny),
    matrix(VV[[1]][,1], nrow = ny, ncol = nx),
    main = "First Latent Spatial Field",
    xlab = "X coordinate", 
    ylab = "Y coordinate"
  )
}
```

## Generate Factor Loadings

Now we'll create factor loadings to link the latent fields to the observed data:

```{r factor_loadings}
# Generate factor loadings matrix
Lambda <- matrix(0, q, actual_k)
diag(Lambda) <- runif(actual_k, 0.5, 1)  # Strong diagonal loadings
Lambda[lower.tri(Lambda)] <- runif(sum(lower.tri(Lambda)), -0.7, 0.7)  # Off-diagonal loadings

# Display the factor loadings matrix
print("Factor loadings matrix (Lambda):")
print(round(Lambda, 2))

# Create covariate matrix (intercept only in this example)
Beta <- matrix(rep(mu, p*q), ncol = q)
print("Beta coefficient matrix:")
print(round(Beta, 2))

# Create design matrices (just intercept)
x_list <- lapply(1:num_images, \(i) {
  matrix(1, nrow = n, ncol = p)
})
```

## Generate Response Data

Finally, we'll generate the observed count data (Poisson responses) based on the latent fields:

```{r generate_response}
# Combine latent spatial fields with factor loadings to create intensity surfaces
WW <- lapply(1:num_images, \(i) {
  mat <- VV[[i]] %*% t(Lambda) + x_list[[i]] %*% Beta
  mat
})

# Generate Poisson counts based on the intensity
y_list <- lapply(WW, \(ww) {
  matrix(rpois(nrow(ww) * ncol(ww), exp(ww)), nrow = nrow(ww), ncol = ncol(ww))
})

# Visualize the first response dimension from the first image
# Create a simple function to plot the response
plot_y <- function(y, coords, dimension = 1, image_num = 1) {
  data <- data.frame(
    x = coords$x * max(x_max, y_max),
    y = coords$y * max(x_max, y_max),
    count = y[[image_num]][, dimension]
  )
  
  ggplot(data, aes(x = x, y = y, fill = count)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme_minimal() +
    labs(
      title = paste("Observed Counts - Image", image_num, "Dimension", dimension),
      x = "X coordinate",
      y = "Y coordinate",
      fill = "Count"
    )
}

# Plot the first dimension of the first image
plot_y(y_list, coords)
```

# Run BIPPS Models

Now we'll run the BIPPS model for each of our trial k values. We'll use the same dataset for all models but vary the number of latent factors:

```{r run_bipps, cache=TRUE}
# Function to run BIPPS for a specific k value
run_bipps_with_k <- function(trial_k) {
  cat("Running BIPPS with k =", trial_k, "\n")
  
  # Run BIPPS
  out <- multi_bipps(
    y_list = y_list,
    x_list = x_list,
    coords = coords,
    k = trial_k,
    family = "poisson",
    block_size = block_size,
    n_samples = n_samples, 
    n_burn = n_burnin, 
    n_thin = n_thin,
    n_threads = n_threads,
    starting = starting,
    prior = prior,
    settings = list(adapting = TRUE, saving = TRUE, ps = TRUE, low_mem = TRUE),
    verbose = 10,
    debug = list(
      sample_beta = TRUE, 
      sample_tausq = FALSE,
      sample_theta = sample_theta, 
      sample_w = TRUE, 
      sample_lambda = TRUE,
      verbose = FALSE, 
      debug = FALSE
    )
  )
  
  return(out)
}

# Run BIPPS for each trial k
results <- list()
for (k in trial_ks) {
  results[[as.character(k)]] <- run_bipps_with_k(k)
}

# Save results
saveRDS(results, "output/bipps_varying_k_results.rds")
```

# Analysis of Results

After running the models, we analyze the results to understand how well different models with varying k values perform. This helps us determine if we can correctly identify the true number of latent factors. We will focus mostly on analysis of the spatial cross-correlations, since these are the most interpretable quantities we get obtain from this model. The spatial cross-correlations are defined:

$$
\rho_{rs}(\mathbf{h}; \mathbf{\theta}) = \frac{C_{rs}(\mathbf{h}; \mathbf{\theta})}{\sqrt{C_{rr}(\mathbf{0}; \mathbf{\theta})} \sqrt{C_{ss}(\mathbf{0}; \mathbf{\theta})}}
$$
$$C_{rs}(\mathbf{h}) = \sum_{j=1}^k \lambda_{rj}\lambda_{sj} \rho(\mathbf{h}; \varphi_j)$$

where $\lambda_{rj}$ is the $(r,j)$ element of the $q \times k$ matrix $\mathbf{\Lambda}$ of factor loadings, $\rho(\mathbf{h}; \varphi_j) = \exp\{ -\varphi_j \mathbf{h} \}$.

## Model Comparison via WAIC

First, let's compare the WAIC (Widely Applicable Information Criterion) scores and compute times across different k values. Lower WAIC scores indicate better model fit.

```{r waic_display}
waic_scores <- unlist(lapply(results,\(o) o$waic))
timings <- unlist(lapply(results,\(o) o$mcmc_time))

# Create data frame for the table
waic_df <- tibble(
  `Number of Latent Factors (k)` = trial_ks,
  `WAIC Score` = waic_scores,
  `Difference from Best` = waic_scores - min(waic_scores),
  `Timing (s)` = timings
)

# Format the table with kable
kableExtra::kable(waic_df, 
      caption = "Model Comparison by WAIC and Computational Time (True k = 2)",
      digits = c(0, 0, 0, 0),
      # format.args = list(big.mark = ","),
      align = c('c', 'r', 'r', 'r')) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "center")

```


## Model Diagnostics

We should also examine MCMC diagnostics to ensure our models have converged properly. Important diagnostics include:

1. Effective Sample Size (ESS)
2. R-hat (Potential Scale Reduction Factor)

First we will extract relevant quantities related to the cross-correlations:

```{r}
# Helper function to get unique combinations including self-pairings
unique_combinations_with_self <- function(data) {
  # Generate all pairs including self-pairings
  all_pairs <- expand.grid(data, data)
  
  # Sort the pairs and remove duplicates
  all_pairs <- t(apply(all_pairs, 1, sort))
  unique_pairs <- unique(all_pairs)
  
  return(unique_pairs)
}

# Distance bins
hs <- seq(0, 1, 0.1)

xl_e <- lapply(trial_ks,\(k) {

  # Extract cross-correlation functions
  xl <- cross_list(results[as.character(k)], hs, thin = n_thin)
  
  # Prepare true cross-correlation values
  # First, reconstruct the "true" model
  out_actual <- list(
    theta_mcmc = matrix(c(philist, rep(0, actual_k)), nrow = 2, ncol = actual_k, byrow = TRUE),
    lambda_mcmc = Lambda
  )
  # Convert to list format expected by cross_list
  out_actual <- list(lapply(out_actual, \(o) {
    dim(o) <- c(dim(o), 1)
    o
  }))
  # Get true cross-correlation values
  xl_actual <- cross_list(out_actual, hs, thin = n_thin)
  xl_actual <- lapply(xl_actual, \(x) E(x))
  
  # Types (response dimensions)
  types <- 1:q
  
  # Calculate cross-correlation metrics for all pairs of types
  xl_e <- unique_combinations_with_self(types) %>%
    as.data.frame() %>%
    as_tibble() %>%
    magrittr::set_colnames(c("t1", "t2")) %>%
    group_by(t1, t2) %>%
    group_modify(~{
      ix1 <- which(types == .y$t1)
      ix2 <- which(types == .y$t2)
      
      # Predicted mean cross-correlation
      mu <- unlist(lapply(xl, \(x) {
        E(x[ix1, ix2])
      }))
      
      # True cross-correlation
      mu_actual <- unlist(lapply(xl_actual, \(x) {
        x[ix1, ix2]
      }))
      
      # Bulk ESS for cross-correlations
      ess_bulk <- as.vector(do.call(rbind,lapply(xl,\(x) {
        posterior::ess_bulk(x[ix1,ix2])
      })))
      
      # Tail ESS for cross-correlations
      ess_tail <- as.vector(do.call(rbind,lapply(xl,\(x) {
        posterior::ess_tail(x[ix1,ix2])
      })))
  
      # Rhat for cross-correlations
      rhats <- as.vector(do.call(rbind,lapply(xl,\(x) {
        posterior::rhat(x[ix1,ix2])
      })))
        
      # Credible intervals
      lb <- unlist(lapply(xl, \(x) {
        quantile(x[ix1, ix2], probs = 0.025)
      }))
      
      ub <- unlist(lapply(xl, \(x) {
        quantile(x[ix1, ix2], probs = 0.975)
      }))
      
      tibble(trial_k = k, actual_k = actual_k, mu = mu, mu_actual = mu_actual,
             lb = lb, ub = ub, hs = hs, rhat = rhats, ess_bulk = ess_bulk,
             ess_tail = ess_tail)
    }) %>%
    ungroup()
}) %>%
  bind_rows()
```

The plots above display three key Markov Chain Monte Carlo (MCMC) diagnostics across different distances and values of k:

### Gelman's Rhat
Rhat (potential scale reduction factor) measures convergence across multiple chains. Values close to 1.0 indicate good convergence, while values substantially above 1.0 (typically >1.01 or >1.05) suggest that chains have not converged to the same posterior distribution.

### Bulk ESS (Effective Sample Size)
Bulk ESS measures the effective number of independent samples for estimating the central part of the posterior distribution. It accounts for autocorrelation in the MCMC chains - higher values indicate more efficient sampling. The bulk ESS is particularly important for accurately estimating posterior means and medians. Low bulk ESS values suggest that despite having many nominal iterations, the effective information content is much lower.

### Tail ESS
Tail ESS focuses specifically on the sampling efficiency in the distribution tails, which is crucial for accurate estimation of quantiles and intervals. Tail ESS is often lower than bulk ESS because extreme values are encountered less frequently during sampling. This diagnostic is particularly important when inference about extreme probabilities or rare events is needed.

```{r}
xl_e %>%
  mutate(trial_k=factor(trial_k)) %>%
  mutate(hs = hs*max(x_max,y_max)) %>%
  group_by(hs,actual_k,trial_k) %>%
  summarise(mean_rhat=mean(rhat,na.rm = T)) %>%
  ggplot(aes(x=hs,y=mean_rhat,color=trial_k,fill=trial_k)) +
  stat_halfeye() +
  labs(y=TeX("Average $\\hat{R}$"),x="Distance (\u03bcm)",color="k",fill="k")

xl_e %>%
  mutate(trial_k=factor(trial_k)) %>%
  mutate(hs = hs*max(x_max,y_max)) %>%
  group_by(hs,actual_k,trial_k) %>%
  summarise(mean_ess=mean(ess_bulk,na.rm = T)) %>%
  ggplot(aes(x=hs,y=mean_ess,color=trial_k,fill=trial_k)) +
  stat_halfeye() +
  labs(y="Average Bulk ESS",x="Distance (\u03bcm)",color="k",fill="k")

xl_e %>%
  mutate(trial_k=factor(trial_k)) %>%
  mutate(hs = hs*max(x_max,y_max)) %>%
  group_by(hs,actual_k,trial_k) %>%
  summarise(mean_ess=mean(ess_tail,na.rm = T)) %>%
  ggplot(aes(x=hs,y=mean_ess,color=trial_k,fill=trial_k)) +
  stat_halfeye() +
  labs(y="Average Tail ESS",x="Distance (\u03bcm)",color="k",fill="k")
```

## MCMC Trace Plots

Trace plots are essential for assessing MCMC convergence. We want to see that the chains have mixed well and are exploring the parameter space efficiently:

```{r trace_plots}
# Get trace plots for spatial decay parameters
if (actual_k %in% trial_ks) {
  # Extract theta (containing phi parameters)
  theta <- get_rvars(results[as.character(actual_k)], "theta", thin = n_thin)
  
  # Plot trace for the first row (phi parameters)
  mcmc_trace(as_draws_df(theta[1, ])) + ggtitle("Trace plot for phi")
}

# Trace plots for factor loadings (Lambda)
lambda <- get_rvars(results[as.character(actual_k)], "lambda", thin = n_thin)
mcmc_trace(as_draws_df(lambda[, 1])) + ggtitle("Trace plot for first column of lambda")  # First column of lambda

# Trace plots for cross-correlations at a specific distance
h_ix <- 5  # Index for distance bin
hs <- seq(0, 1, 0.1)  # Distance bins
xl <- cross_list(results[as.character(actual_k)], hs, thin = n_thin)

# Create and plot trace data
trace_df <- as_draws_df(xl[[h_ix]]) %>%
  pivot_longer(-c(".chain", ".iteration", ".draw"), names_to = "variable") %>%
  separate(variable, into = c("type1", "type2"), sep = ",") %>%
  mutate(
    type1 = sub("^x\\[", "", type1),
    type2 = sub("\\]", "", type2)
  )

# Center each trace around its mean to better visualize mixing
ggplot(
  trace_df %>%
    mutate(.chain = factor(.chain)) %>%
    mutate(type1 = factor(type1, levels = 1:q)) %>%
    mutate(type2 = factor(type2, levels = 1:q)) %>%
    group_by(type1, type2) %>%
    mutate(value = value - mean(value)) %>%
    ungroup(),
  aes(.iteration, value)
) +
  geom_line() +
  facet_grid(
    type1 ~ type2,
    labeller = label_wrap_gen(width = 8)
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x = "Draw",
    y = paste0("Mean-centered cross-correlation at h=", hs[h_ix] * max(x_max, y_max), "μm")
  )
```

Here we can see that our MCMC chains definitely don't seem to have converged, which makes sense, since we only had 500 burn-in samples! Running for longer should make convergence look much better.

# Analyzing Spatial Cross-Correlation

One of the key features of BIPPS is its ability to model the spatial cross-correlation between different types (or components) of the response. Here, we'll analyze how well our models capture these spatial relationships.

## Spatial Cross-Correlation Analysis

We'll compute and visualize the spatial cross-correlation functions for different pairs of response types:

```{r cross_correlation}
# Plot cross-correlations
xl_e %>%
  filter(trial_k == actual_k) %>%
  mutate(hs = hs * max(x_max, y_max)) %>%
  rename(Predicted = mu, True = mu_actual) %>%
  pivot_longer(c(Predicted, True)) %>%
  ggplot() +
  geom_ribbon(aes(hs, ymin = lb, ymax = ub), fill = "grey70") +
  geom_line(aes(hs, value, color = name)) +
  geom_hline(yintercept = 0, color = "red", alpha = 0.5, linetype = "dotted") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  facet_grid(t1 ~ t2) +
  labs(
    x = "Distance (μm)",
    y = "Cross-correlation",
    color = "Cross-correlation"
  )
```

The plot above shows the spatial cross-correlation functions for each pair of response types. For each pair:

- The blue line represents the true cross-correlation from our model
- The red line represents the estimated cross-correlation function
- The grey ribbon shows the 95% credible interval for our estimate
- The x-axis represents the spatial distance in micrometers

When our model is performing well, the estimated cross-correlation should closely follow the true cross-correlation, and the credible intervals should include the true values.

## Comparing Intensity Surfaces

We can also compare the estimated intensity surfaces with the true ones and the observed counts:

```{r intensity_comparison}
# Plot observed counts, true intensity, and predicted intensity
# Select a subset of images and types for visualization
image_idx <- 1
types_idx <- 1:2

# Extract observed counts
y_list_subset <- lapply(y_list[image_idx], \(yy) {
  yy <- yy[, types_idx]
  colnames(yy) <- paste0("Cell type: ", types_idx)
  yy
})

# Extract true intensity surfaces
WW_subset <- lapply(WW[image_idx], \(ww) {
  ww <- ww[, types_idx]
  colnames(ww) <- paste0("Cell type: ", types_idx)
  ww
})

# Create plots
py <- plot_y_list(y_list_subset, coords * max(x_max, y_max))
pW <- plot_y_list(WW_subset, coords * max(x_max, y_max))

# Get predicted intensity
# Extract posterior means for V (latent factors) and Lambda (loadings)
vhat <- results[[as.character(actual_k)]]$v_mcmc
vhat <- lapply(vhat,\(v) v[1:(nx*ny),])
arr <- simplify2array(vhat)
arr <- aperm(arr,perm=c(3,1,2))
vhat <- posterior::rvar(arr)

beta <- results[[as.character(actual_k)]]$beta_mcmc
arr <- beta
arr <- arr[,,seq(1,dim(arr)[3],by=n_thin),drop=FALSE]
arr <- aperm(arr,perm=c(3,1,2))
beta <- posterior::rvar(arr)

# Calculate predicted intensity surface
lp_e <- E(vhat %*% t(lambda) + matrix(1,nrow=nx*ny,ncol=p) %*% beta)
colnames(lp_e) <- paste0("Cell type: ", 1:q)

# Plot predicted intensity
p3 <- dplyr::bind_cols(lp_e[,types_idx],coords * max(x_max,y_max)) %>%
  tidyr::pivot_longer(-c(x,y),names_to = "type",values_to = "count") %>%
  ggplot2::ggplot(ggplot2::aes(x,y,fill=count)) +
  ggplot2::geom_tile() +
  ggplot2::facet_wrap(~type,ncol=2) +
  ggplot2::scale_fill_viridis_c(option="magma") +
  theme(axis.text = element_blank(),
        axis.title.y=element_blank()) +
  guides(fill="none")


py[[1]]
pW[[1]] + labs(fill="log-intensity")
p3
```

The figures shown are:

- Plot 1: Observed count data for selected cell types
- Plot 2: True underlying intensity that generated the counts
- Plot 3: Model-predicted intensity surfaces

When the model fits well, the predicted intensity surfaces should closely match the true intensity surfaces, despite the randomness in the observed counts.

# Conclusion

In this example, we demonstrated:

1. How to set up and run BIPPS simulations with varying numbers of latent factors
2. The process of generating synthetic spatial point process data
3. How to fit BIPPS models with different specifications
4. Methods for comparing model performance and evaluating results

The simulation framework allows us to test how well BIPPS can recover the true underlying structure of spatial point process data. By comparing models with different numbers of latent factors, we can develop insights into model selection and specification.

## Next Steps

Potential extensions of this analysis could include:

1. Varying other parameters like the spatial decay range
2. Testing with different point process families (e.g., Negative Binomial)
3. Adding covariates to the model
4. Exploring different spatial correlation structures
